Place of death in NRS COVID-19 deaths

### 2021-02-18 covid_nrs_place_of_death.Rmd
started incorporating covid case counts, scope for lots of calculations; need to normalise health board names between datasets

---
title: "Place of death in NRS COVID-19 deaths"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
params:
  data_sources: '`r tibble()`'
  historical_data_sources: '`r tibble()`'
---

@@

```{r}
endpoint <- "http://statistics.gov.scot/sparql.json"
```

@@

```{r}
# "unpack" the data source tables
table_of_data_sources <- params$data_sources
table_of_historical_data_sources <- params$historical_data_sources

@@

# ods_all_datasets() %>% View()
# ods_structure("deaths")  # can we import all deaths-related data this way?
# ods_print_query("deaths")

@@

```{r}
weekly_deaths_overall %>% 
  count(ref_period, date_w_c, week_number, week_number_run_over) %>% 
  slice(1:3, (nrow(.) - 3):nrow(.)) %>% 
  select(-n) %>% 
  knitr::kable(caption = "Checking the computation of dates from week numbers - note how the 1st week of 2020 started in 2019.")
```

@@

@@

# Summarise annual deaths for 2020 & 2021

>>>>

---
title: "Place of death in NRS COVID-19 deaths"
author: "Jan Savinc"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
params:
  data_sources: '`r tibble()`'
  historical_data_sources: '`r tibble()`'
  case_trends_data_sources: '`r tibble()`'
---

@@

```{r}
endpoint <- "http://statistics.gov.scot/sparql.json"
endpoint <- "http://statistics.gov.scot/sparql"
```

@@

```{r}
# "unpack" the data source tables
table_of_data_sources <- params$data_sources
table_of_historical_data_sources <- params$historical_data_sources
table_of_case_trends_data_sources <- params$case_trends_data_sources

@@

ods_all_datasets() %>% View()
ods_structure("deaths")  # can we import all deaths-related data this way?
ods_structure("coronavirus-covid-19-management-information")

ods_dataset("coronavirus-covid-19-management-information", geography = "hb") %>%
  filter(variable == "testing-new-cases-reported")

ods_dataset("coronavirus-covid-19-management-information", geography = "hb", variable = "testing-new-cases-reported")
ods_dataset("coronavirus-covid-19-management-information", geography = "la", variable = "testing-new-cases-reported")
ods_dataset("coronavirus-covid-19-management-information", variable = "testing-new-cases-reported")

ods_print_query(dataset = "coronavirus-covid-19-management-information", geography = "hb", variable = "testing-new-cases-reported") %>% dput
ods_print_query(dataset = "coronavirus-covid-19-management-information", variable = "testing-new-cases-reported") %>% dput

modified_query <- "PREFIX qb: <http://purl.org/linked-data/cube#>                 PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>                 PREFIX xsd: <http://www.w3.org/2001/XMLSchema#> select   ?refArea ?refPeriod ?measureType ?variable ?value where { ?data qb:dataSet <http://statistics.gov.scot/data/coronavirus-covid-19-management-information>. ?data <http://purl.org/linked-data/sdmx/2009/dimension#refArea> ?refArea. ?data <http://purl.org/linked-data/sdmx/2009/dimension#refPeriod> ?refPeriod. ?data <http://purl.org/linked-data/cube#measureType> ?measureType. ?data <http://statistics.gov.scot/def/dimension/variable> ?variable. ?data ?measureType ?value.  filter (regex(str(?variable ), 'testing-new-cases-reported'))  } order by ?refPeriod ?refArea"

modified_query %>%
  opendatascot:::ods_query_database(endpoint, query = .) %>%
  opendatascot:::pre_process_data()

query <- ""

library(ckanr)
ckanr_setup(url = "https://www.opendata.nhs.scot/")
# package_list(as = "table")
# tag_list(as = "table")
package_search(q = 'daily', rows = 2, as = 'table') %>% str

download_data_from_opendata_via_ckan_api <- function(resource_id) {
  ckan <- ckanr::src_ckan("https://www.opendata.nhs.scot/")
  dplyr::tbl(src = ckan$con, from = resource_id_daily_case_trends_la, ) %>% 
  as_tibble()
}

resource_id_daily_case_trends_la <- "427f9a25-db22-4014-a3bc-893b68243055"

resource_show(id = resource_id_daily_case_trends_la, as = "table")

# ods_print_query("deaths")

@@

```{r}
weekly_deaths_sex_age %>% 
  count(year, date_w_c, week_number, week_number_run_over) %>% 
  slice(1:3, (nrow(.) - 3):nrow(.)) %>% 
  select(-n) %>% 
  knitr::kable(caption = "Checking the computation of dates from week numbers - note how the 1st week of 2020 started in 2019.")
```

@@

## Load COVID case counts by LA/HB

```{r}
files_case_counts <- set_names(x = as.list(table_of_case_trends_data_sources$path_latest), nm = table_of_case_trends_data_sources$short_name)

cases_hb <- read_csv(
  files_case_counts$hb, 
  col_types = 
    cols(
    .default = col_double(),
    `_id` = col_skip(),
    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
    Date = col_date(format = "%Y%m%d"),
    HB = col_character(),
    HBName = col_character(),
    HospitalAdmissionsQF = col_character(),
    ICUAdmissionsQF = col_character()
    ),
  ) %>% clean_names() %>%
  rename(ref_area = hb_name) %>%
  mutate(
    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8))
    )

cases_la <- read_csv(
  file = files_case_counts$la, 
  col_types = 
    cols(
    .default = col_double(),
    `_id` = col_skip(),
    DailyDeaths = col_skip(),  # not needed, we get this elsewhere!
    Date = col_date(format = "%Y%m%d"),
    CA = col_character(),
    CAName = col_character()
  )
) %>% clean_names() %>%
  rename(ref_area = ca_name) %>%
  mutate(
    year = parse_integer(str_sub(ISOweek(date), start = 1, end = 4)),  # parse ISO year, otherwise the first few days of 2021 will belong to 2021 erroneously!
    week_number = parse_integer(str_sub(ISOweek(date), start = 7, end = 8))
    )
```

# Calculate cumulative case counts at weekly deaths

Note: the cumulative case counts have already been calculated, we just need to subset them to match the weeks. This is not trivial - the death counts cover an entire week starting from the "w/c date"; the cumulative cases for that week could be taken either at the start, the end of the week, or the start of the following week even!
However, when calculating weekly positive cases we can simply take the maximum of that week's cumulative count - that will be the cumulative count at the end of the week!

```{r}
weekly_covid_cases_hb <-
  cases_hb %>%
  group_by(year, week_number, ref_area) %>%
  summarise(
    weekly_positive = sum(daily_positive),
    cumulative_positive = max(cumulative_positive),
    .groups = "drop"
  ) %>%
  mutate(
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
    )

```

@@

# Annual deaths for 2020 & 2021
